## Extending the Logistic Model

## Case Studies

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Age and Heart Attacks (fake data, I made it up)

Heart_data <- read_csv(file =           'https://raw.githubusercontent.com/gitcnk/Data/master/Heart_Attacks.csv')

HA <- ggplot() +
        geom_jitter(data = Heart_data, 
                    aes(x = Age, y = Heart_Attack, col = as.factor(Heart_Attack) ),                          height = 0.05,
                    alpha = 0.3) +
        scale_color_discrete(name='Status', labels = c('None','Heart Attack')) +
        labs(title = 'Risk of Heart Attacks ',
             subtitle = 'Sample of Patients with High Cholesterol',
             y = 'Probability of Heart Attack')
HA

```



```{r message=FALSE, warning=FALSE, include=FALSE}
status_table <- table( Heart_data$Heart_Attack, Heart_data$group )

status_proportions <- prop.table(status_table, margin = 2)

fit <- locpoly( x = Heart_data$Age, y = Heart_data$Heart_Attack, bandwidth = 1.3)


prop_data <- data.frame(x = seq(33,65, length.out = 31), y = status_proportions[2, ])
fit_data <- data.frame(fit)

```

Plot with local averages

```{r echo=FALSE, message=FALSE, warning=FALSE}
HA + geom_point( data = prop_data, aes(x = x, y = y)) 
```


Plot with the smoothed curve.
```{r echo=FALSE, message=FALSE, warning=FALSE}
HA + geom_point( data = prop_data, aes(x = x, y = y)) +
     geom_line( data = fit_data, aes(x = x, y = y), col = 'red') 
```




**Me:** Also, there is another issue.  When you have more than one $X$ variables, then you can not visualize the data and the estimates in a plot.  For example, suppose you what to model the probability using 3 variables.  Then it is impossible to visualize the data.  Do you know how to draw a 4D plot?

**You:**  Why 4D?  You said we 3 $X$ variables, so why cant we use a 3D plot?

**Me:**  Don't forget the response variable, $Y$, hence the $4^{th}$ dimension.  If I push it even further, what if you have 10 $X$ variables, then you need a 11D plot!!

**You** I see what you mean.

**Me:** We encountered this same issue in the previous chapter on multiple regression models.








```{r echo=FALSE, message=FALSE, warning=FALSE}
library(Sleuth3)
ford_tire_data <- ex2018
ford_tire_data$status_code <- recode(ford_tire_data$Cause, 'Tire' = 1, 'NotTire' = 0)
str(ford_tire_data) # display the STRUCTURE of a data frame

ford <- ford_tire_data %>% filter(Make =='Ford')
other <- ford_tire_data %>% filter(Make !='Ford')



fordplot <- ggplot() +
            geom_jitter(data = ford,
                        aes( x = VehicleAge, y = status_code, col = Cause),
                        height = 0.1 ) +
            labs(title = 'FORD') +
            theme(legend.position = "bottom")



t <- table(ford$VehicleAge, ford$Cause)
tt <- prop.table(t, margin = 1)
ford_prop_data <- data.frame( x = seq(0,5), y = tt[,2])

p1 <- fordplot + geom_point(data = ford_prop_data, aes(x = x, y = y))
p1

otherplot <- ggplot() +
                geom_jitter(data = other,
                            aes( x = VehicleAge, y = status_code, col = Cause),
                            height = 0.1 ) +
                labs(title = 'OTHER')  +
                theme(legend.position = "none")


t <- table(other$VehicleAge, other$Cause)
tt <- prop.table(t, margin = 1)
other_prop_data <- data.frame( x = seq(0,5), y = tt[,2])

p2 <- otherplot + geom_point(data = other_prop_data, aes(x = x, y = y))
p2


ggplot() + 
  geom_point(data = ford_prop_data, aes(x = x, y = y), col = 'red') +
  geom_point(data = other_prop_data, aes(x = x, y = y), col = 'blue') +
  labs(title = 'The Risk of Accidents Linked to Tires' ,
       subtitle = 'Ford (red dots) vs. Others (blue dots)', 
       x = 'Vehicle Age',
       y = 'Proportion of Accidents')
  
```


 Predicted risk using the Logistic model.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ford_tire_data) +
  aes(x = VehicleAge, y = status_code, col = Make) +
  geom_jitter(width = 0.1,
              height = 0.1) +
  geom_smooth(method = 'glm', 
              method.args = list(family = 'binomial'),
              se = FALSE, 
              fullrange = TRUE)




m1 <- glm(Cause ~ VehicleAge + Make + Passengers, 
         family = 'binomial',
         data = ford_tire_data)
#summary(m1)
kable(tidy(m1), format = 'html', digits = 3)

m2 <- glm(Cause ~ VehicleAge  + Passengers, 
         family = 'binomial',
         data = ford_tire_data)
#summary(m2)
kable(tidy(m2), format = 'html', digits = 3)


logLik(m1)
logLik(m2)



```



